Somatic misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C

### 2024-06-29 submission/manuscript.Rmd
drafting intro

---
title:  Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C
output: pdf_document
geometry: margin = 0.5in
font-size: 11pt
header-includes:
   - \usepackage{caption}
   - \captionsetup[figure]{font=footnotesize, labelfont=bf}
   - \usepackage{sectsty} 
   - \usepackage{helvet}
   - \renewcommand{\familydefault}{\sfdefault}
   - \usepackage{lineno}
   - \linenumbers
   - \renewcommand\linenumberfont{\normalfont\small}
   - \usepackage{setspace}
   - \spacing{1.5}
   - \pagenumbering{arabic}

@@

>>>>

---
title:  Somatic misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C
output: pdf_document
geometry: margin = 0.5in
font-size: 11pt
header-includes:
   - \usepackage{caption}
   - \captionsetup[figure]{font=footnotesize, labelfont=bf}
   - \usepackage{sectsty} 
   - \usepackage{helvet}
   - \renewcommand{\familydefault}{\sfdefault}
   - \usepackage{lineno}
   - \linenumbers
   - \renewcommand\linenumberfont{\normalfont\small}
   - \usepackage{setspace}
   - \spacing{1.5}
   - \pagenumbering{arabic}
   - \usepackage{indentfirst}

@@

# Introduction

To form a complete organism, embryonic stem cells must differentiate into a myriad of discrete cellular identities. This is in part accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through histone and DNA modifications. Although initially identified for their roles in cellular identity [@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing technologies unexpectedly found many neurodevelopmental disorders (NDDs) are caused by or linked to mutations in chromatin regulators. This relationship is partially explained by their regulation of brain-specific genes or chromatin states, such as modulating genes involved in synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] or the transition between neuronal and glial developmental programs[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also lead to the misexpression of tissue-specific genes outside of their intended environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Currently, very few studies have explored the misexpression of non-neuronal, tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is unclear if this partial loss of brain identity contributes to neurodevelopmental impairments. To elucidate their contribution to neurodevelopmental impairments, it is essential to first characterize the types of genes misexpressed, the developmental time point the dysregulation begins, and the molecular mechanism underlying their de-repression. Characterizing these features will provide the basis for identifying molecular footholds that can then be exploited for future studies and potential therapeutics.

In this study, we characterized the misexpression of tissue-specific with loss of the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose features include short stature, intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Previous work has demonstrated _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression and learning impairments[@iwaseMouseModelXlinked2016]. Next generation RNA sequencing (RNA-seq) in the _Kdm5c_-KO hippocampus surprisingly revealed ectopic expression of testis-enriched genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is currently unclear if misexpression in the _Kdm5c_-KO brain is unique to male germline genes, as other tissue-specific genes have not been systematically evaluated.

Interestingly, some of the ectopic testis genes identified in the _Kdm5c_-KO brain are typically expressed in germ cells, such as _Cytochrome C, Testis-specific_, a sperm mitochondrial gene[cit].

- Germ cells are meiotic cells that pass on their genetic material to the next generation and are typically distinguished from somatic cells very early on in embryogenesis. 
- The demarcation of the germ vs soma is a key feature of multicellularity

* Germ cells, i.e. cells that undergo meiosis, are typically distinguished from somatic cells very early on in embryogenesis and this distinction is a key feature of multicellularity.

* However, the testis also contains somatic cells that support germline function, and it is currently unclear if  It is currently unclear if all testis- (maybe too specific)

	* Chromatin regulators are very important for decommissioning germline genes and act successively the embryo implants into the uterine wall
		* Most studies have focused on ESCs, which have a similar transcriptome to germ cells / 2-cells
		* recently, KDM5C was shown to repress DAZL in ESCs, independent of its catalytic activity
		* However, DNA methylation is lost in the mature 5CKO brain, DNA methylation is placed later and it's Unclear if it's required for long-term repression (maybe too specific, just trying to go into the fact that the mechanism is partially understood but unclear)

Furthermore, the testis contains both somatic and germline

It is currently unclear if these germline genes

	* Systematic characterization of ectopic germline genes hasn't been done
		* unknown if other germline-enriched genes are dysregulated, including oocyte-specific genes
		* Crucially, it's unknown if misexpression of the germline program leads to functional consequences in 5CKO cells.

However, it is

Germ cells (meiotic cells) are typically distinguished from somatic cells very early on in embyrogenesis and is a key feature of multicellularity

	* Misexpression of tissue-enriched genes hasn't been systematically characterized - Unclear if these genes are exceptions or if other tissue-specific genes are dysregulated

	* Interestingly, these genes (Cyct, D1pas1) typically function in the germline
	* Germ cells (meiotic cells) are typically distinguished from somatic cells very early on in embyrogenesis and is a key feature of multicellularity
	* Chromatin regulators are very important for decommissioning germline genes and act successively the embryo implants into the uterine wall
		* Most studies have focused on ESCs, which have a similar transcriptome to germ cells / 2-cells
		* recently, KDM5C was shown to repress DAZL in ESCs, independent of its catalytic activity
		* However, DNA methylation is lost in the mature 5CKO brain, DNA methylation is placed later and it's Unclear if it's required for long-term repression (maybe too specific, just trying to go into the fact that the mechanism is partially understood but unclear)

Answering these questions will help identify molecular footholds that can be exploited to test their overall contribution to neuronal impairments that will lay the ground work for therapeutic intervention.

Loss of KDM5C can result in the misexpression of genes typically only found in the testis
	* Misexpression of tissue-enriched genes hasn't been systematically characterized - Unclear if these genes are exceptions or if other tissue-specific genes are dysregulated
	* Interestingly, these genes (Cyct, D1pas1) typically function in the germline
	* Germ cells (meiotic cells) are typically distinguished from somatic cells very early on in embyrogenesis and is a key feature of multicellularity
	* Chromatin regulators are very important for decommissioning germline genes and act successively the embryo implants into the uterine wall
		* Most studies have focused on ESCs, which have a similar transcriptome to germ cells / 2-cells
		* recently, KDM5C was shown to repress DAZL in ESCs, independent of its catalytic activity
		* However, DNA methylation is lost in the mature 5CKO brain, DNA methylation is placed later and it's Unclear if it's required for long-term repression (maybe too specific, just trying to go into the fact that the mechanism is partially understood but unclear)
	* Systematic characterization of ectopic germline genes hasn't been done
		* unknown if other germline-enriched genes are dysregulated, including oocyte-specific genes
		* Crucially, it's unknown if misexpression of the germline program leads to functional consequences in 5CKO cells.

To determine the

- To ultimately determine their contribution, it is essential to characterize the types of genes expressed and the mechanism behind their de-repression.

	* Necessary to first characterize the mechanism behind their derepression to identify molecular footholds into testing their contribution to neuronal impairments and potential for therapeutic intervention

	* However, loss of some chromatin regulators can also lead to ectopic expression of tissue-enriched genes
		* Very few studies have looked at these genes and it's unclear if these genes contribute to NDD impairments. 
	* Necessary to first characterize the mechanism behind their derepression to identify molecular footholds into testing their contribution to neuronal impairments and potential for therapeutic intervention

- For example, histone 3 lysine 4 methylation (H3K4me)

- [@lugerCrystalStructureNucleosome1997].

- This relationship is partially 
It is currently unc

- There is a strong link between H3K4me regulation and NDDs, as mutation of 11 out of 15 genes that read, write, or erase H3K4 mono-, di-, and tri- methylation result in varying neurodevelopmental impairments and syndromic NDDs[@vallianatosDisruptedIntricacyHistone2015]. Yet it is still unclear why regulation of H3K4 methylation is so crucial for neurodevelopment, as most of H3K4 methylation regulators are ubiquitously expressed across tissues, cell types, and lifespan.

With the recent advancement of next generation sequencing technologies came the unexpected finding that many neurodevelopmental disorders are caused by or linked to mutations in genes that regulate chromatin[@derubeisSynapticTranscriptionalChromatin2014; @gabrieleChromatinBasisNeurodevelopmental2018; @iwaseEpigeneticEtiologyIntellectual2017a].

Chromatin regulators, including those that  DNA and histone modification, is an integral component of cellular identity

Chromatin regulators, including those that  DNA and histone modification, is an integral component of cellular identity

Cellular identity 
Some of the first chromatin regulators , such as

With the recent advancement of next generation sequencing technologies came the unexpected finding that many neurodevelopmental disorders are caused by or linked to mutations in genes that regulate chromatin[@derubeisSynapticTranscriptionalChromatin2014; @gabrieleChromatinBasisNeurodevelopmental2018; @iwaseEpigeneticEtiologyIntellectual2017a].

The basic subunit of chromatin is the nucleosome and is comprised of DNA wrapped around octamers of histone proteins[@lugerCrystalStructureNucleosome1997].

Histone proteins can undergo covalent chemical modifications on their N-terminal tails that can in turn influence cofactor recruitment, DNA accessibility, and gene expression[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001].
